Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
Updates
restructuring

\title{Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge}

@@

\begin{enumerate}
    \item We develop an iterative method based on CI testing to assist researchers in 
	    constructing DAGs using their domain expertise (Section~\ref{sec:modification}).
    \item We show a ranking method to prioritize potential modifications to the DAG that 
    	    help in fixing the most severe violations first (Section~\ref{sec:ranking}).
    \item We provide a web-based interactive tool and a Python package to allow 
	    researchers to easily apply this method to their own datasets (Section~\ref{sec:web}).
    \item Lastly, we compare the manual DAG construction method with automated
	    causal discovery algorithms (Section~\ref{sec:empirical}).
\end{enumerate}

\section{Background and Related Work}
\label{sec:background}
\todo[inline]{Need to rewrite this}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, $ \mathrm{corr}(X, Y) $ is the correlation between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.  Covariance matrix using $ \Sigma $ and the entry corresponding to
covariance between $ X $ and $ Y $ in the matrix using $ \Sigma_{XY} $.

\subsection{CI Tests and Measures of Conditional Association}
Our structure learning approach relies on using CI tests and a measure of
conditional association between variables -- also known as partial association.
Specifically, we are interested in testing CI tests: $ X \ci Y \bm{Z} $, i.e.,
whether the variables $ X $ and $ Y $ are independent when conditioned on $
\bm{Z} $. Secondly, we are interested in quantifying the strength of
relationship between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We can use
the effect size measures of the CI tests for this quantification. In this 
section, we give an overview of some of the commonly used CI tests for 
various data types and their effect size measures.

@@

\paragraph{All $ X $, $ Y $, and $ \bm{Z} $ are discrete: }

When all variables are discrete, Cram\'er's V can used as a measure of
association. When $ \bm{Z} = \emptyset $, Cram\'er's V is derived from the 
chi-squared test statistic:

\begin{equation}
	\mathbf{V}_{X, Y} = \sqrt{\frac{\chi^2 / n}{\mathrm{min}(k-1, r-1)}}
\end{equation}

where $ \chi^2 $ is the chi-squared test statistic for $ X $ and $ Y $, $ n $
is the sample size, and $ k $ and $ r $ are the number of categories in $ X $
and $ Y $ respectively. When $ \bm{Z} \neq \emptyset $, we start by splitting
the dataset by each unique combination of $ \bm{Z} $ into subsets,  $ \bm{Z} $,
$ D = \{ D_{\bm{Z} = \bm{Z}_1}, D_{\bm{Z} = \bm{Z}_2}, \cdots, D_{\bm{Z} =
\bm{Z}_k} \} $. Then we compute the Cram\'er's V for each of these smaller datasets
and combine them as follows:

\begin{equation}
	\mathbf{V}_{X, Y; \bm{Z}} = \sum_{i=1}^{k} \left[ \mathbf{V}_{X, Y} \right]_{D_{\bm{Z} = \bm{Z}_i}} 
\end{equation}

@@

 The implied CIs of the 
model along with the test results are shown in Figure.

\section{DAG Structure Learning using Ancestral Oracles}
\label{sec:modification}

@@

@@

To achieve this, instead of selecting edges randomly, we propose ranking
potential edges based on their residual association given the current DAG. This
residual association can be quantified using the effect size measures from the
CI tests used for querying the d-separation in Algorithm~\ref{algo:expand}.
Specifically, for a d-separation query $ D_G(X, Y, \mathbf{Z}) $, we quantify
the residual association, $ \phi(X, Y, \bm{Z}) $, as the effect size of the CI
test $ X \ci Y \rvert \bm{Z} $. In Algorithm~\ref{algo:ranked_expand}, we
outline a greedy approach to edge selection using the effect size measure. The
algorithm selects the edge that has the highest absolute residual association,
resulting in prioritization of edges that contribute the most to improving the
model. We use this \textsc{RankedExpand} in all of our experiments in the
following sections.

\subsection{Comparison to Automated Algorithms}
\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\textsc{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\textsc{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
		$\alpha_{\textrm{eff}} $ is shown in the plot.}
	\label{fig:shd_sid}
\end{figure}

@@

\subsection{Connection to Score Based Methods}

@@

Another key difference lies in the interpretability of the evaluation metric.
Scoring metrics are based on usually based on the log-likelihood with a penalty
for model complexity. They can be used to make relative comparisons between
models but their values do not have any interpretation in an absolute sense.
That is, they indicate which model is better for a given dataset but do not
quantify how well the model explains the data in an absolute sense. In
contrast, $ \tau $ provides an absolute measure of model fit -- its value
approaches $ 0 $ as the model perfectly explains the observed data.
Figure~\ref{fig:unexplained_ll} illustrates this distinction by comparing GES
to \textsc{ExpertInLoop} on the Adult income dataset across iterations of a
single run of the algorithms.

@@

Since our approach relies on expert knowledge to determine edge orientations,
we explored the potential of using an LLM for this task. We applied our causal
discovery procedure using the Gemini 1.5 Flash model as the expert. We simulate
the \emph{weak ancestral oracle} using the LLM by asking it to choose the
causal direction between a pair of variables. The LLM is provided a description
of each of the variables for the task. We have provided the prompt used in
Appendix~\ref{section:llms}. Figure~\ref{fig:adult_llm} shows the DAG learned
on the Adult Income dataset using the LLM as the expert. Compared to other
methods that use LLMs for causal discovery, our approach combines a data-driven
method with LLMs. We believe this should help in reducing the amount of
mistakes made by the LLM as well as reduce the amount of information needed
from the LLM.

\section{Web Tool}
\label{sec:web}

>>>>

\title{Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge}

@@

\begin{enumerate}
    \item We develop an iterative method based on CI testing to assist researchers in 
	    constructing DAGs using their domain expertise (Section~\ref{sec:modification}).
    \item We show a ranking method to prioritize potential modifications to the DAG that 
    	    help in fixing the most severe violations first (Section~\ref{sec:ranking}).
    \item We show empirically that even experts that make mistakes can outperform purely data-driven
	causal discovery algorithms (Section~\ref{sec:empirical}).
    \item We provide two proof-of-concept implementations of our approach: one geared towards
	LLMs as experts, and a graphical user interface designed 
	for human experts (Section~\ref{sec:web}).
\end{enumerate}

\section{Background}
\label{sec:background}
\todo[inline]{Need to rewrite this}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, $ \mathrm{corr}(X, Y) $ is the correlation between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.  Covariance matrix using $ \Sigma $ and the entry corresponding to
covariance between $ X $ and $ Y $ in the matrix using $ \Sigma_{XY} $.

 The implied CIs of the 
model along with the test results are shown in the Figure.

We can think of our proposed approach to be an iterative application of this model
testing approach.

\subsection{Effect Sizes for CI Tests}

Our structure learning approach relies on using CI tests and a measure of
conditional association between variables -- also known as partial association.
Specifically, we are interested in testing CI tests: $ X \ci Y \bm{Z} $, i.e.,
whether the variables $ X $ and $ Y $ are independent when conditioned on $
\bm{Z} $. Secondly, we are interested in quantifying the strength of
relationship between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We can use
the effect size measures of the CI tests for this quantification. In this 
section, we give an overview of some of the effect size measures for 
different types of (possibly mixed) data.

@@

@@

\paragraph{$ X $, $ Y $, and $ \bm{Z} $ are discrete (ordinal or caterical): }

\todo[inline]{Here we use an effect size RMSEA based on the previous Hotelling's test paper.}

\section{Expert-In-The-Loop Causal Discovery}

@@

We will now first present theoretical results showing how a structure
learning algorithm using such experts can uncover the true DAG structure.
Afterwards, we will present a heuristic for deciding which changes should
be prioritized.

\subsection{DAG Structure Learning using Ancestral Oracles}

\label{sec:modification}

@@

To achieve this, instead of selecting edges randomly, we propose ranking
potential edges based on their residual association given the current DAG. This
residual association can be quantified using the effect size measures from the
CI tests used for querying the d-separation in Algorithm~\ref{algo:expand}.
Specifically, for a d-separation query $ D_G(X, Y, \mathbf{Z}) $, we quantify
the residual association, $ \phi(X, Y, \bm{Z}) $, as the effect size of the CI
test $ X \ci Y \rvert \bm{Z} $. In Algorithm~\ref{algo:ranked_expand}, we
outline a greedy approach to edge selection using the effect size measure.

To implement Algorithm~\ref{algo:ranked_expand}, we need to have an effect size 
metric by which we can rank effects. For some simple cases, such as 
linear-Gaussian models, such effect sizes exist. However, a challenge arises
when using mixed data. We here present a standardized effect size hat 
is useable for datasets containing mixtures of continuous, ordinal and

The
algorithm selects the edge that has the highest absolute residual association,
resulting in prioritization of edges that contribute the most to improving the
model.

\subsection{Connection to Score-Based Methods}

Another key difference lies in the interpretability of the evaluation metric.
Scoring metrics are based on usually based on the log-likelihood with a penalty
for model complexity. They can be used to make relative comparisons between
models but their values do not have any interpretation in an absolute sense.
That is, they indicate which model is better for a given dataset but do not
quantify how well the model explains the data in an absolute sense. In
contrast, $ \tau $ provides an absolute measure of model fit -- its value
approaches $ 0 $ as the model perfectly explains the observed data.

In all experiments, we use the \textsc{RankedExpand} method together
with a version of the strong ancestral oracle that can make mistakes.

\todo{Want to know: what if my experts are not perfect?}

\begin{figure}[t!]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC, Hill-Climb Search, and GES algorithms against
		\textsc{ExpertInLoop} algorithm. As automated algorithms only
		recover the CPDAG, we use the best and worst scoring
		orientation of the CPDAG to get the range. We test
		\textsc{ExpertInLoop} with varying values of expert accuracy, $ \alpha = \{0.1, 0.3, 0.5, 0.7, 0.9\} $. The corresponding
		$\alpha_{\textrm{eff}} $ is shown in the plot.}
	\label{fig:shd_sid}
\end{figure}

@@

@@

\section{Practical Implementation}

\label{sec:web}

We now present two proof-of-concept implementations of our approach, 
geared towards using  large language models (LLMs) and humans as experts
to decide on edge directions.

@@

Since our approach relies on expert knowledge to determine edge orientations,
we explored the potential of using an LLM for this task. We applied our causal
discovery procedure using the Gemini 1.5 Flash model as the expert. We simulate
the \emph{weak ancestral oracle} using the LLM by asking it to choose the
causal direction between a pair of variables. The LLM is provided a description
of each of the variables for the task. We have provided the prompt used in
Appendix~\ref{section:llms}. 
Figure~\ref{fig:unexplained_ll} illustrates this distinction by comparing GES
to \textsc{ExpertInLoop} on the Adult income dataset across iterations of a
single run of the algorithms.
Figure~\ref{fig:adult_llm} shows the DAG learned
on the Adult Income dataset using the LLM as the expert. Compared to other
methods that use LLMs for causal discovery, our approach combines a data-driven
method with LLMs. We believe this should help in reducing the amount of
mistakes made by the LLM as well as reduce the amount of information needed
from the LLM.

\subsection{Using Humans as Experts}
